---
title: "Juvenile Chinook age estimation"
author: "mark sorel"
date: "8/18/2020"
output:
  pdf_document: default
  html_document: default
---

The purpose of this document is to explore how to determine the ages of juvenile Chinook captured in screw traps in the  Wenatchee River Basin.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

knitr::opts_chunk$set(tidy = FALSE, highlight = TRUE, comment = NA, 
                      dev = "png", dev.args = list(type = "cairo-png"), dpi = 200,
                      out.width = "75%", fig.align = "center")
```

```{r Read_biodata, echo=FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(readxl)
library(mixtools)
#The first step is to read in the data with lengths and dates of individual fish captured
#CHIWAWA DATA
chiw_fry_biod<-read_excel(here("data","Chiwawa","originals","RSTChiwawaSpringChinookTotalCaptures1997_2018_new.xlsx"),sheet=5) %>% mutate(DOY=as.numeric(format(DATE,form="%j")),year=as.numeric(format(DATE,form="%Y"))) #fry

chiw_sub_biod<-read_excel(here("data","Chiwawa","originals","RSTChiwawaSpringChinookTotalCaptures1997_2018_new.xlsx"),sheet=6)%>% mutate(DOY=as.numeric(format(DATE,form="%j")),year=as.numeric(format(DATE,form="%Y"))) #parr

chiw_smolt_biod<-read_excel(here("data","Chiwawa","originals","RSTChiwawaSpringChinookTotalCaptures1997_2018_new.xlsx"),sheet=7)%>% mutate(DOY=as.numeric(format(DATE,form="%j")),year=as.numeric(format(DATE,form="%Y"))) #yearlings

chiw_biod<-rbind(chiw_fry_biod,chiw_sub_biod,chiw_smolt_biod) %>% rename(Length=`FORK (mm)`) #all combined


#NASON and WHITE
 #read in biological tagging data
  bioD<-read.csv(here("data","Nason and White","Compiled Biodata.csv")) %>% 
  
  #change a column name that reads in wrong
  rename(Trap=1) %>% 
  #add a column of stream name
  mutate(stream=ifelse(substr(Trap,1,1)=="N","Nason","White"))%>% 
  
  #supstet to wild spring Chinook
  subset(Species=="Wild Spring Chinook") %>% 
  
  #drop precocial males and adults
  subset(Stage!="A"&Stage!="PR") %>% 
  
  #add a reformated "Date" column
  mutate(Date2=as.Date(Date,format="%m/%d/%Y"),
  #add DOY
  DOY=as.numeric(format(Date2,format="%j")),
  
  #make "length" numeric
  Length=as.numeric(as.character(Length))) %>% droplevels()
  
  #subsetting by river
  wSpC_Nason<-subset(bioD,stream=="Nason",)
  wSpC_White<-subset(bioD,stream=="White",)

#combining all rivers
  all_bio_data<- full_join(bioD,chiw_biod,by=c("Length","DOY")) %>% 
    filter(Length>=20 & Length<=200)                  #filter 3 outliers
```  


Histogram of capture day-of-year across all streams
```{r , echo=FALSE, warning=FALSE}
hist(all_bio_data$DOY,breaks=seq(50,365,5),xlab="Day of year",main="All traps")

# test<-mixtools::normalmixEM(
#     na.exclude(all_bio_data$DOY),
#     lambda=c(.4,.2,.4),
#     mu=c(100,200,300),
#     sigma=c(10,10,10))
# plot(test,2)
# summary(test)
```
  


```{r read_PTAGIS, echo=FALSE, warning=FALSE}
#read data on fish that were tagged at screw traps and subsequently detected at mainstem dams when migrating downstream as smolts. This provides information on dates and lengths of fish that were known to be smolts or subyearlings.

#mark sites
mark_site_codes<-c("CHIWAT","NASONC","WHITER")

#juvenile mainstem dam detectin sites
juv_mainst<-c("MCJ","JDJ","BCC","B2J","TWX","B1J")

#load and munge interrogation data
interrogation<-read.csv(here("data","PTAGIS","wen wild tag inter hist.csv")) %>% subset(Mark.Site.Info.Code %in% mark_site_codes & Site.Code.Value %in% juv_mainst & Mark.Capture.Method.Code=="SCREWT") %>% arrange_at("First.Year.YYYY") %>% subset(!duplicated(Tag.Code)) %>% mutate(Mark_to_mig=First.Year.YYYY-Mark.Year.YYYY)
```

Histogram of capture day of year of known smolts (fish that were tagged at screw traps and subsequently detected at mainstem dams when migrating downstream, in the same year they were tagged) 

The 99.9th quantile of capture days-of-year for know smolts (n =`r length(interrogation[interrogation$Mark_to_mig==0,"Release.Day.Number"])`)  is **`r quantile(interrogation[interrogation$Mark_to_mig==0,"Release.Day.Number"],.999) `**
```{r , echo=FALSE, warning=FALSE}
hist(interrogation[interrogation$Mark_to_mig==0,"Mark.Day.Number"],breaks=seq(50,365,5),xlab="Day of year",main="Known smolts")
smolt_max_DOY<-quantile(interrogation[interrogation$Mark_to_mig==0,"Release.Day.Number"],.999)
abline(v=smolt_max_DOY)
```


Fish captured on or before day 
**`r quantile(interrogation[interrogation$Mark_to_mig==0,"Release.Day.Number"],.999) `**  

  
```{r , echo=FALSE, warning=FALSE}
 spring_all_bio_data<-subset(all_bio_data,DOY<=smolt_max_DOY & !is.na(Length)) %>% select(Length,DOY) # subsetting to spring

 #plot(spring_all_bio_data$DOY,spring_all_bio_data$Length,ylab="Length (mm)",xlab="Day of year",col=rgb(.5,.5,.5,.5),pch=19,cex=.4)                     #plotting
```


To create a line to separate subyearlings and yearlings:
1) fit a mixture distribution of two normal distributions to the log(length) data for each 10-day interval starting at day 50, 
2) find the length corresponding to the minimum density between the two modes of the mixture distribution,
and 3) connect the points.

```{r eval=TRUE, message=FALSE, warning=FALSE,echo=F,out.width="40%",fig.show="hold"}

knots<-seq(50,179,by=10)              #starting days for 10-day intervals
n_knots<-length(knots)                #number of 10-day intervals
break_ln<-numeric(length(n_knots))    #vector to hold break points for each interval


for ( i in 1:n_knots){                #loop through intervals 
  x<-subset(all_bio_data,DOY>=knots[i]
            & DOY<(knots[i]+10))      #subset to interval
  
 hist(x[,"Length"],xlab="Length",xlim=c(20,140),
     breaks=seq(0,200,2),freq = FALSE,main=paste0("Days ",knots[i],"-",knots[i]+9))#plot histogram
 
  
   invisible(capture.output( mixd<-mixtools::normalmixEM(
    na.exclude(log(x$Length)),
    lambda=.5,
    mu=c(3.5,4.5),
    sigma=c(1,1))))                    #fit mixture distribution of 2 normals
   
   
  
int_seq<-seq(exp(mixd$mu[1]),
             exp(mixd$mu[2]),by=.1)  # sequence of lengths between modes of mixture components to "predict" to
                                      
mix_pred<-mixd$lambda[1]*dlnorm(int_seq,mixd$mu[1],mixd$sigma[1])+
mixd$lambda[2]*dlnorm(int_seq,mixd$mu[2],mixd$sigma[2])#mixture density at .1 mm intervals

break_ln[i]<- int_seq[which.min(    # find length between modes with minimum density
mix_pred)]

abline(v=break_ln[i])               #add breakpoint to histogram  

int_seq<-(20:140)             
points(int_seq,mixd$lambda[1]*dlnorm(int_seq,mixd$mu[1],mixd$sigma[1])+
mixd$lambda[2]*dlnorm(int_seq,mixd$mu[2],mixd$sigma[2]),type="l",col="red") #add mixture distribution to histogram

}

```

Add line to plot of fish captured on or before day 
**`r quantile(interrogation[interrogation$Mark_to_mig==0,"Release.Day.Number"],.999) `**
```{r , echo=FALSE, warning=FALSE}
 
 plot(spring_all_bio_data$DOY,spring_all_bio_data$Length,ylab="Length (mm)",xlab="Day of year",col=rgb(.5,.5,.5,.5),pch=19,cex=.4)                      # plotting
 
#calculate the line end points (day 50 and 179)
first_point_y<-((break_ln[2]-break_ln[1])/10*-5)+break_ln[1]
last_point_y <-((break_ln[n_knots]-break_ln[n_knots-1])/10*4)+break_ln[n_knots]

points(c(50,knots+5,smolt_max_DOY),c(first_point_y,break_ln,last_point_y),type="l",col="red") 

daily_cutoff<-approx(c(50,knots+5,smolt_max_DOY),c(first_point_y,break_ln,last_point_y),xout = 50:179)

#calculate proportions of each age on each DOY, for approtioning fish with no length mesurements.

daily_proportions<-list(x=daily_cutoff$x,prop=numeric(length(daily_cutoff$x)))
for ( i in daily_proportions$x){
  sub_all<-which(all_bio_data$DOY==i)
    if(length(sub_all)>0){
      daily_proportions$prop[(i-49)]<-sum(all_bio_data$Length[sub_all]>=daily_cutoff$y[c(i-49)],na.rm=T) /length(sub_all)
    }
}

props_lo<-loess(daily_proportions$prop~daily_proportions$x)
props_lo_pred<-predict(props_lo,data.frame(x=50:179))
props_lo_pred[props_lo_pred<0]<-0
daily_proportions$prop_lo_pred<-props_lo_pred

cutoffs_and_props<-list(daily_cutoff,daily_proportions)
save(cutoffs_and_props,file=here("data","processed","cutoffs_and_props.Rdata"))
```



Now overlay fish known to be subyearling and yearlings based on whether they were detected at mainstem dams when migrating downstream in the same year they were tagged or the next year.

```{r , echo=FALSE, warning=FALSE}
 
 plot(spring_all_bio_data$DOY,spring_all_bio_data$Length,ylab="Length (mm)",xlab="Day of year",col=rgb(.5,.5,.5,.5),pch=19,cex=.4)                      # plotting
 
points(c(50,knots+5,smolt_max_DOY),c(first_point_y,break_ln,last_point_y),type="l",col="red")

  points(interrogation$Mark.Day.Number,interrogation$Mark.Length.mm,col=ifelse(interrogation$Mark_to_mig==0,"purple","darkgreen"),cex=.75) 
```


```{r , echo=FALSE, warning=FALSE}
 
 plot(all_bio_data$DOY,all_bio_data$Length,ylab="Length (mm)",xlab="Day of year",col=rgb(.5,.5,.5,.5),pch=19,cex=.4)                      # plotting
 
#calculate the line end points (day 50 and 179)
points(c(50,knots+5,smolt_max_DOY,smolt_max_DOY),c(first_point_y,break_ln,last_point_y,200),type="l",col="red",lwd=1.5)

  points(interrogation$Mark.Day.Number,interrogation$Mark.Length.mm,col=ifelse(interrogation$Mark_to_mig==0,"purple","darkgreen"),cex=.75) 
```


Not bad

```{r}
load(here("data","processed","cutoffs_and_props.Rdata"))
interrogation<-filter(interrogation,!is.na(Mark.Length.mm)) %>% mutate(pred_Mark_to_mig=ifelse(Mark.Day.Number>179,1,ifelse(Mark.Length.mm>=cutoffs_and_props[[1]]$y[(Mark.Day.Number-49)],0,1)))

tab<- table(interrogation$pred_Mark_to_mig,interrogation$Mark_to_mig)


thing<-interrogation %>% filter(pred_Mark_to_mig!=Mark_to_mig) %>% select(Tag.Code,Mark.Site.Info.Code,Mark.Date.MMDDYYYY,Mark.Year.YYYY,Mark.Day.Number,Mark.Length.mm,Mark_to_mig,pred_Mark_to_mig,Site.Name,First.Date.MMDDYYYY,First.Year.YYYY,First.Day.Num)
write.csv(thing,file=here("data","processed","misassigned_fish.csv"))

```

The assignment of "known age" fish n=(`r sum(tab)`) is correct in `r round((tab[1,1]+tab[2,2])*100/sum(tab),2)`% of cases 

```{r,eval=F,echo=FALSE}
#attempting to optimize a piecewise linear function with maximum distance to points. Doesn,t really work

knots<-seq(50,180,by=10)
n_knots<-length(knots)
xout<-50:80

foo<-function(par,data_in){

x<-approx(knots,par,xout=50:180)$y

obj<-0
for ( i in xout){
  sub_dat<-subset(data_in,DOY==i)
  if(nrow(sub_dat)>0){
obj<-obj+sum(abs(sub_dat[,"Length"]-x[(i-49)]))    
  }
}
return(-obj)
}


test_optim<-optim(par=runif(n_knots,60,75),foo,lower=rep(43,n_knots),upper=rep(75,n_knots),data_in=spring_all_bio_data,method="L-BFGS-B")
test_optim

plot(50:180,approx(knots,test_optim$par,xout=50:180)$y)

```



```{r plot, echo=FALSE, warning=FALSE,eval=FALSE}  
#relic old plotting code for individual screw traps

 #Chiwawa

{plot(chiw_fry_biod$DOY, chiw_fry_biod$`FORK (mm)`,xlim = c(50,360),ylim=c(0,200),col=rgb(.9,.1,.1,.3),pch=19,cex=.4,ylab="Length",xlab='DOY',main="Chiwawa River")
points(chiw_sub_biod$DOY, chiw_sub_biod$`FORK (mm)`,col=rgb(.9,.1,.1,.3),pch=19,cex=.4)
points(chiw_smolt_biod$DOY, chiw_smolt_biod$`FORK (mm)`,col=rgb(.1,.1,.9,.3),pch=19,cex=.4)
}

{plot(chiw_fry_biod$DOY, chiw_fry_biod$`FORK (mm)`,xlim = c(50,225),ylim=c(0,200),col=rgb(.9,.1,.1,.3),pch=19,cex=.4,ylab="Length",xlab='DOY',main="Chiwawa River")
abline(v=179)
points(chiw_sub_biod$DOY, chiw_sub_biod$`FORK (mm)`,col=rgb(.9,.1,.1,.3),pch=19,cex=.4)
points(chiw_smolt_biod$DOY, chiw_smolt_biod$`FORK (mm)`,col=rgb(.1,.1,.9,.3),pch=19,cex=.4)
  
segments(c(0,110,200),c(55,55,100),c(110,200,350),
           c(55,100,137.5),col="black")

int_chwiawa<-subset(interrogation,Mark.Site.Code.Value=="CHIWAT")
  points(interrogation$Mark.Day.Number,interrogation$Mark.Length.mm,col=ifelse(int_chwiawa$Mark_to_mig==0,"purple","yellow"),cex=.75) 


}
  
   #Nason Creek
{points(wSpC_Nason$DOY,as.numeric(as.character(wSpC_Nason$Length)),pch=19,cex=.4,type="p",col=rgb(.1,.1,.1,1),ylab="Length",xlab='DOY',main="Nason Creek")
  
  #this line is what I came up with, where everythng below is a subyearling and everything above is a yearling
  segments(c(0,110,200),c(55,55,100),c(110,200,350),
           c(55,100,137.5),col="red")
  
  #abline(h=50,col="red")
    int_nason<-subset(interrogation,Mark.Site.Code.Value=="NASONC")
  points(int_nason$Mark.Day.Number,int_nason$Mark.Length.mm,col=ifelse(int_nason$Mark_to_mig==0,"blue","red"))  

}

  #White River
{ points(wSpC_White$DOY,as.numeric(as.character(wSpC_White$Length)),pch=19,cex=.4,type="p",col="green",ylab="Length",xlab='DOY',main="White River")
  
 
   #this line is what I came up with, where everythng below is a subyearling and everything above is a yearling
  segments(c(0,110,200),c(55,55,100),c(110,200,350),
           c(55,100,137.5),col="red")
 # abline(h=50,col="red")
  
      int_white<-subset(interrogation,Mark.Site.Code.Value=="WHITER")
  points(int_white$Mark.Day.Number,int_white$Mark.Length.mm,col=ifelse(int_white$Mark_to_mig==0,"blue","red"))  
  
 }
 
  
```
